Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-18 ebayes_rnaseq_heterosis.tex
added map estimates

We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau_\gamma^2)$. (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and take $\tau_\gamma\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution. In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$. In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

@@

We found MAP estimates for both the independence and covariance models as shown in Table \ref{t:map}. 
% Need MAP estimates table and discussion

>>>>

We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau^2)$. (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and take $\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution. In the independent model, we also assign the $Ca^+(0,3)$ prior to the diagonal elements of $\Sigma$. In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

@@

We found MAP estimates for both the independent and covariance models as shown in Table \ref{t:map}. 
\input{include/map}
The MAP estimates for the hierarchical expression means, $\eta_i = E[\mu_{gi}]$ for $i=1,2,3$, are consistent between the the independent and covariance model while the mean for the overdispersion parameter, $\eta_4=E[\log(\psi)]$, is a bit large in the covariance model. The standard deviation around the expression means are approximately twice as large in the covariance model as in the independent model while the standard deviation for the overdispersion is half as large. The sequencing depth terms are consistent between the two models and indicate that there is about a 25\% increase in counts between the least and most thoroughly sequenced samples. In the covariance model, there appears to be moderate correlation amongst the expression means and the expression means are negatively correlated with the (log) overdispersion parameter.
